Categories of Optics

### 2018-04-09 paper.tex
Progress with diagrams

\usepackage{tikz-cd}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{cd}
\usetikzlibrary{arrows}
\input{tikzcurverel}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
%\newtheorem{axiom}[theorem]{Axiom}
\newtheorem*{theoremstar}{Theorem}
%\newtheorem{fact}[theorem]{Fact}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

@@

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Lawful}{\mathbf{Lawful}}
\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Tamb}{\mathbf{Tamb}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

@@

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
&\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
 \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact, and the Fubini theorem for coends \cite[Remark 1.9]{CoendCofriend}, the domain is isomorphic to
\begin{align*}
\int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps 
\begin{align*}
& \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
 \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
&\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
\to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
\to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
\to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\langle l, r \rangle : (R, R') \hto (S, S')$ and $\langle l', r' \rangle : (S, S') \hto (A, A')$ with $M$ the complement for $\langle l, r \rangle$. The composite is then: $\langle l', r' \rangle \circ \langle l, r \rangle = \langle (M l')l, r(Mr')\rangle$.

Graphically, this is placing the second optic into the middle of the first:
\begin{center}
\todo{todo}
\end{center}

The identity optic $(S, S') \hto (S, S')$ is given by $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$.

@@

We check associativity of composition equationally. As in the definition of composition, the Fubini theorem allows us to choose representatives for all three optics simultaneously. Suppose we have representatives of three optics $\langle l_1, r_1\rangle : (R, R) \hto (S, S')$ and $\langle l_2, r_2 \rangle : (S, S') \hto (A, A')$ and $\langle l_3, r_3 \rangle : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. Then:
\begin{align*}
(\langle l_3, r_3 \rangle \circ \langle l_2, r_2 \rangle) \circ \langle l_1, r_1 \rangle 
&= \langle (N \otimes l_3)l_2, r_2(N \otimes r_3) \rangle \circ \langle l_1, r_1 \rangle \\
&= \langle (M \otimes ((N \otimes l_3)l_2))l_1, r_1(M \otimes (r_2(N \otimes r_3))) \rangle \\
&= \langle (M \otimes N \otimes l_3)(M \otimes l_2)l_1, r_1(M \otimes r_2)(M \otimes N \otimes r_3)) \rangle \\
&= \langle l_3, r_3 \rangle \circ (\langle (M \otimes l_2)l_1, r_1(M \otimes r_2) \rangle) \\
&= \langle l_3, r_3 \rangle \circ (\langle l_2, r_2 \rangle \circ \langle l_1, r_1 \rangle)
\end{align*}
Graphically:
\begin{center}
\todo{todo}
\end{center}

For the unit laws, suppose we have $\langle l, r \rangle : (S, S') \hto (A, A')$ with representative $M$. We calculate:
\begin{align*}
\id_{A, A'} \circ \langle l, r\rangle 
&= \langle \lambda^{-1}_A, \lambda_{A'} \rangle \circ \langle l, r\rangle \\
&= \langle (M \otimes \lambda^{-1}_A) l, r (M\otimes  \lambda_{A'})\rangle \\
&= \langle (\lambda^{-1}_M \otimes  A) l, r (\lambda_M \otimes A')\rangle \\
&= \langle l, r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')\rangle \\
&= \langle l, r \rangle  \\
\langle l, r \rangle \circ \id_{S, S'} 
&= \langle l, r \rangle \circ \langle \lambda^{-1}_S, \lambda_{S'}\rangle  \\
&= \langle (I \otimes l)\lambda^{-1}_S, \lambda_{S'} (I \otimes r) \rangle \\
&= \langle (\lambda^{-1}_M \otimes S)l, r (\lambda_{M} \otimes S') \rangle \\
&= \langle l, r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S') \rangle \\
&= \langle l, r \rangle
\end{align*}
Where in both cases we have used the naturality of $\lambda$ and coend relations to move $\lambda^{-1}_M$ to the right-hand side of the coend. Graphically:
\begin{center}
\todo{todo}
\end{center}

%<---
\end{proof}

@@

\begin{proposition}
There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
\iota(f', g') \circ \iota(f, g) 
&= \langle \lambda^{-1}_B f', g' \lambda_{B'} \rangle \circ \langle \lambda^{-1}_A f, g \lambda_{A'} \rangle \\
&= \langle (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f, g \lambda_{A'} (I\otimes (g' \lambda_{B'}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f, g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'})\rangle && \text{(Functoriality of $I \otimes -$)}\\
&= \langle (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (I\otimes \lambda_{B'})\rangle && \text{(Naturality of $\lambda$)}\\
&= \langle (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B')\rangle && \text{(Unitality of action)} \\
&= \langle \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B') \rangle && \text{(Coend relation)}  \\
&= \langle \lambda^{-1}_B f'f, g g' \lambda_{B'} \rangle \\
&= \iota(f'f, gg')
\end{align*}
Graphically, there is not much to do:
\begin{center}
\todo{todo}
\end{center}
\end{proof}

@@

\begin{theorem}
$\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, T' \otimes S')$, the unit is $(I, I)$, and the action on morphisms is given by:
\begin{center}
\input{diagrams/tensor-on-morphisms.tikz}
\end{center}
\end{theorem}
\begin{proof}
Equationally, suppose we have representatives $(l, r) : (S \to M \otimes A, M \otimes A' \to S')$ and $(l', r') : (T \to N \otimes B, N \otimes B' \to T')$. Then:
\begin{align*}
\langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
\end{align*}
This is functorial, as suggested by the equivalence of the following string diagrams:
\todo{todo}
%For future convenience, note that 
%\begin{align*}
%\langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
%&= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
%&= \langle l \otimes T, r \otimes T' \rangle
%\end{align*}
%Now we check functoriality. Suppose we have:
%\begin{align*}
%\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
%\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
%\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
%\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
%\end{align*}
%with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
%\begin{align*}
%&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
%= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
%= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
%= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
%= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
%\end{align*}

The structure morphisms are all lifted from the structure morphisms in $\C$:
\begin{align*}
\alpha_{(R, R'), (S, S'), (T, T')} &:= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1}) \\
\lambda_{(S, S')} &:= \iota(\lambda_{(S, S')}, \lambda_{(S, S')}^{-1}) \\
\rho_{(S, S')} &:= (\rho_{(S, S')}, \rho_{(S, S')}^{-1}) \\
s_{(S, S'), (T, T')} &:= \iota(s_{S, T}, s_{T', S'})
\end{align*}

@@

In the Haskell \texttt{lens} library, the action of the monoidal product on optics is denoted ``\texttt{alongside}'' for the product and ``\texttt{without}'' for the coproduct.

\begin{proposition}
$\iota : \C \times \C^\op \to \Optic_\C$ is a strong monoidal functor. \qed
\end{proposition}

@@

@@

\subsection{Teleological Categories}

As noted in \cite{CoherenceForLenses}, \todo{the category of optics admits its own string diagram language}

\begin{definition}[{\cite[Definition 5.1]{CoherenceForLenses}}]
A \emph{teleological category is} a symmetric monoidal category $(\C, \otimes, I)$, equipped with:
\begin{itemize}
\item A wide symmetric monoidal subcategory $\C_d$ of \emph{dualisable morphisms}, with an involutive strong symmetric monoidal functor $(-)^* : \C_d \to \overline{\C_d^\op}$, where---not finding a standard symbol for such a thing---we mean $\overline{\C_d^\op}$ to be the category with both the direction of the arrows \emph{and} the order of the tensor flipped: $(A \otimes B)^* = B^* \otimes A^*$.
\item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\C_d$.
\end{itemize}
\end{definition}
Note that because $\C_d$ is symmetric monoidal and has the same collection of objects as $\C$, the symmetric monoidal structure of $\C$ must consist of dualisable maps.

>>>>

\usepackage{tikz-cd}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{cd}
\usetikzlibrary{arrows}
\tikzset{ed/.style={auto,inner sep=2pt,font=\scriptsize}} %edges
\tikzset{>=stealth'}
\tikzset{vert/.style={draw,circle, minimum size=6mm, inner sep=0pt, fill=white}}
\tikzset{vertbig/.style={draw,circle, minimum size=8mm, inner sep=0pt, fill=white}}
\input{tikzcurverel}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[subsection]
%\newtheorem{axiom}[theorem]{Axiom}
\newtheorem*{theoremstar}{Theorem}
%\newtheorem{fact}[theorem]{Fact}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

@@

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\Lawful}{\mathbf{Lawful}}
\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Setter}{\mathbf{Setter}}
\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Tamb}{\mathbf{Tamb}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

@@

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
&\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
 \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact, and the Fubini theorem for coends \cite[Remark 1.9]{CoendCofriend}, the domain is isomorphic to
\begin{align*}
\int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps 
\begin{align*}
& \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
 \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
&\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
\to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
\to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
\to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\langle l', r' \rangle : (R, R') \hto (S, S')$ and $\langle l, r \rangle : (S, S') \hto (A, A')$ with $M$ the complement for $\langle l', r' \rangle$. The composite is then: $\langle l, r \rangle \circ \langle l', r' \rangle = \langle (M l)l', r'(Mr)\rangle$.

Graphically, this is placing the second optic into the middle of the first:
\begin{center}
\input{diagrams/generic-optic-noline.tikz}
\qquad\raisebox{1.5cm}{$\circ$}\qquad
\input{diagrams/generic-optic2-noline.tikz} \\
\raisebox{1.5cm}{$:=$}\qquad
\input{diagrams/optic-composition.tikz}
\end{center}

The identity optic $(S, S') \hto (S, S')$ is given by $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$: 
\begin{center}
\input{diagrams/identity-optic-full.tikz}
\end{center}
This dashed line above the diagram represents the unit object. As is standard, we will elide the unit object and unitors in future diagrams. The diagram for the identity becomes simply:
\begin{center}
\input{diagrams/identity-optic.tikz}
\end{center}

@@

We check associativity of composition equationally. As in the definition of composition, the Fubini theorem for coends allows us to choose representatives for all three optics simultaneously. Suppose we have representatives of three optics $\langle l_1, r_1\rangle : (R, R) \hto (S, S')$ and $\langle l_2, r_2 \rangle : (S, S') \hto (A, A')$ and $\langle l_3, r_3 \rangle : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. Then:
\begin{align*}
(\langle l_3, r_3 \rangle \circ \langle l_2, r_2 \rangle) \circ \langle l_1, r_1 \rangle 
&= \langle (N \otimes l_3)l_2, r_2(N \otimes r_3) \rangle \circ \langle l_1, r_1 \rangle \\
&= \langle (M \otimes ((N \otimes l_3)l_2))l_1, r_1(M \otimes (r_2(N \otimes r_3))) \rangle \\
&= \langle (M \otimes N \otimes l_3)(M \otimes l_2)l_1, r_1(M \otimes r_2)(M \otimes N \otimes r_3)) \rangle \\
&= \langle l_3, r_3 \rangle \circ (\langle (M \otimes l_2)l_1, r_1(M \otimes r_2) \rangle) \\
&= \langle l_3, r_3 \rangle \circ (\langle l_2, r_2 \rangle \circ \langle l_1, r_1 \rangle)
\end{align*}

For the unit laws, suppose we have $\langle l, r \rangle : (S, S') \hto (A, A')$ with representative $M$. We calculate:
\begin{align*}
\id_{A, A'} \circ \langle l, r\rangle 
&= \langle \lambda^{-1}_A, \lambda_{A'} \rangle \circ \langle l, r\rangle \\
&= \langle (M \otimes \lambda^{-1}_A) l, r (M\otimes  \lambda_{A'})\rangle \\
&= \langle (\lambda^{-1}_M \otimes  A) l, r (\lambda_M \otimes A')\rangle \\
&= \langle l, r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')\rangle \\
&= \langle l, r \rangle  \\
\langle l, r \rangle \circ \id_{S, S'} 
&= \langle l, r \rangle \circ \langle \lambda^{-1}_S, \lambda_{S'}\rangle  \\
&= \langle (I \otimes l)\lambda^{-1}_S, \lambda_{S'} (I \otimes r) \rangle \\
&= \langle (\lambda^{-1}_M \otimes S)l, r (\lambda_{M} \otimes S') \rangle \\
&= \langle l, r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S') \rangle \\
&= \langle l, r \rangle
\end{align*}
Where in both cases we have used the naturality of $\lambda$ and coend relations to move $\lambda^{-1}_M$ to the right-hand side of the coend. 
\end{proof}

@@

\begin{proposition}
There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
Graphically, this is:
\begin{center}
\input{diagrams/iota.tikz}
\end{center}

This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
\iota(f', g') \circ \iota(f, g) 
&= \langle \lambda^{-1}_B f', g' \lambda_{B'} \rangle \circ \langle \lambda^{-1}_A f, g \lambda_{A'} \rangle \\
&= \langle (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f, g \lambda_{A'} (I\otimes (g' \lambda_{B'}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f, g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'})\rangle && \text{(Functoriality of $I \otimes -$)}\\
&= \langle (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (I\otimes \lambda_{B'})\rangle && \text{(Naturality of $\lambda$)}\\
&= \langle (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B')\rangle && \text{(Unitality of action)} \\
&= \langle \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B') \rangle && \text{(Coend relation)}  \\
&= \langle \lambda^{-1}_B f'f, g g' \lambda_{B'} \rangle \\
&= \iota(f'f, gg')
\end{align*}
Graphically, there is not much to do:
\begin{center}
\input{diagrams/iota-functorial-left.tikz}
\qquad \raisebox{0.3cm}{$=$} \qquad 
\input{diagrams/iota-functorial-right.tikz}
\end{center}

\end{proof}

@@

\begin{theorem}
$\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit is $(I, I)$, and the action on morphisms is given by:
\begin{center}
\input{diagrams/tensor-on-morphisms.tikz}
\end{center}
\end{theorem}
\begin{proof}
Equationally, suppose we have optics $S \hto A$ and $T \hto B$ with representatives:
\begin{align*}
(l, r) &: (S \to M \otimes A, M \otimes A' \to S') \\
(l', r') &: (T \to N \otimes B, N \otimes B' \to T')
\end{align*}
Then:
\begin{align*}
\langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
\end{align*}
This is functorial, as suggested by the equivalence of the following string diagrams:
\todo{todo}
%For future convenience, note that 
%\begin{align*}
%\langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
%&= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
%&= \langle l \otimes T, r \otimes T' \rangle
%\end{align*}
%Now we check functoriality. Suppose we have:
%\begin{align*}
%\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
%\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
%\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
%\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
%\end{align*}
%with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
%\begin{align*}
%&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
%= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
%= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
%= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
%= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
%\end{align*}

The structure morphisms are all lifted from the structure morphisms in $\C$:
\begin{align*}
\alpha_{(R, R'), (S, S'), (T, T')} &:= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1}) \\
\lambda_{(S, S')} &:= \iota(\lambda_{(S, S')}, \lambda_{(S, S')}^{-1}) \\
\rho_{(S, S')} &:= \iota(\rho_{(S, S')}, \rho_{(S, S')}^{-1}) \\
s_{(S, S'), (T, T')} &:= \iota(s_{S, T}, s_{T', S'})
\end{align*}

@@

\begin{remark}
In the Haskell \texttt{lens} library, the monoidal product on optics is denoted ``\texttt{alongside}'' for the product and ``\texttt{without}'' for the coproduct.
\end{remark}

\begin{proposition}
If $\C$ is a strict symmetric monoidal category then $\Optic_\C$ is also strict.
\end{proposition}
\begin{proof}
The structure maps of $\Optic_\C$ are given by $\iota$ applied to the structure maps of $\C$. If the latter are identities, then so are the former---the identity morphisms in $\Optic_\C$ are by definition $\iota(\id_S, \id_S')$.
\end{proof}

\begin{proposition}
$\iota : \C \times \C^\op \to \Optic_\C$ is a strong monoidal functor. 
\end{proposition}
\begin{proposition}
\todo{todo}
\end{proposition}

@@

In particular, for any $M \in \C$, the image of the identity yields an optic $c_M : (M, M) \hto (I, I)$, that we will call the connector.

@@

\subsection{Teleological Categories}
\label{teleological-categories}

Categories of optics admit their own diagrammatic language, by virtue of them being \emph{teleological categories}. In fact, we show in Theorem \ref{optic-is-free-teleological-cat} that $\Optic_\C$ is the free teleological category on a collection of dualisable morphisms $\C$.

In this section we assume that $\C$ is a \emph{strict} symmetric monoidal category.
